Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-06-21 analysis/paper/paper.qmd
editing

<!--summary of overall observations across time periods -->
The sampled area at Kiwulan contains the large quantity of faunal remains from the European period, followed by the late pre-European period. Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but only half the quantity. This might indicate a consistent pattern of regular deer consumption at the site both before and after European presence, with intensified deer use during the European period. In addition to the dominance of deer, the increased relative proportion and ubiquity of suids in the European period suggests a more regular use of suids compared to other periods. The dramatic decline in deer bones after the European period may be associated with both a reduction in the Kiwulan population and the decrease of wild deer. Sika deer populations in western Taiwan largely declined from the 18th century due to habitat loss and overhunting [@Chen2000animal]. A slight increase in muntjac use suggests that they may have served as a substitute, replacing suids as a secondary animal resource from earlier periods. Despite these shifts, deer remained a primary resource for local consumption at Kiwulan during the post-European and early Chinese periods. Bovines were present in small numbers (yellow cattle) in the pre-European period, but became the sole large animal (water buffalo) used at Kiwulan in the late Chinese period. Since bovines were absent from prehistoric sites in Taiwan and are widely believed to have been introduced from mainland China in historical period [@Chen2000animal], their increasing presence suggests Chinese influence and aligns with the Indigenous relocation to eastern Taiwan since the 19th century due to Chinese population pressures [@Zhan1995].

>>>>

<!--summary of overall observations across time periods -->
The sampled area at Kiwulan contains the large quantity of faunal remains from the European period, followed by the late pre-European period. Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but only half the quantity. This might indicate a consistent pattern of regular deer consumption at the site both before and after European presence, with intensified deer use during the European period. In addition to the dominance of deer, the increased relative proportion and ubiquity of suids in the European period suggests a more regular use of suids compared to other periods. The dramatic decline in deer bones after the European period may be associated with both a reduction in the Kiwulan population and the decrease of wild deer. Sika deer populations in western Taiwan largely declined from the 18th century due to habitat loss and overhunting [@Chen2000animal]. A slight increase in muntjac use suggests that they may have served as a substitute, replacing suids as a secondary animal resource from earlier periods. Despite these shifts, deer remained a primary resource for local consumption at Kiwulan during the post-European and early Chinese periods.

Bovines were present in small numbers (yellow cattle) in the pre-European period but became the only animal (water buffalo) in the late Chinese period (CL1). One possible explanation for the shift in the animal taxonomic structure from deer to bovines is the dramatic decline of the Indigenous Kavalan population in Yilan during the 19th century, around the same time as a large influx of Han Chinese immigrants. Competition over land between the two groups eventually led to several waves of Kavalan relocation to eastern Taiwan, leaving only a small number of Kavalan in Yilan [@Zhan1995]. Furthermore, bovines were absent in prehistoric Taiwan and are believed to have been introduced by Han populations during the historical period [@Chen2000animal]. The reliance on bovines in CL1 may indicate both a decline in the Indigenous population and a shift from traditional hunting to livestock use introduced or intensified under Han influence. A similar cultural change is also observed in burial practices at Kiwulan, where coffin-like containers, typical funerary traditions of Han Chinese, were first adopted by Indigenous people in the late period (CL1 and CL2) [@Chen2007; @Chiu2004].
